### 2023-01-27 paper.Rmd
some 'last changes'

[^dissimilarityfromfeatures]: It is still possible to convert the attribute data into a dissimilarity matrix, e.g. by computing all pairwise Euclidean distances, and then optimize a distance-based anticlustering criterion such as the diversity to obtain a partitioning.

@@

Because the sample variance is basically defined as an average---and k-means anticlustering minimizes differences with regard to averages---k-means anticlustering can be employed to equalize the variance across groups. This constitutes the central insight motivating the new k-plus criterion.

To obtain an adaptation of the $\mathit{SSE}$ that quantifies the degree to which the variance differs between groups, a new variable is computed for each original variable; the new variable represents the squared distance of each of the original values to the overall mean of the variable. We obtain a new data matrix $\mathbf{Y}_{N \times P} = \{y_{ij}\}_{N \times P}$ where $y_{ij} = (x_{ij} - \overline{\mathbf{x}}_j)^2$. On the basis of the data matrix $\mathbf{Y}_{N \times P}$ we define the criterion $\mathit{SSE_{Var}}$, reflecting how strongly the variances of all attributes vary across the $K$ clusters:

@@

This formula again has the same form as $\mathit{SSE}$ but is computed on different data. In this case, both the original values $x_{ij}$ as well as the squared distances $y_{ij}$ are used as attributes that enter the objective function. That is, computing the k-plus criterion can be done by first computing the values $y_{ij}$, appending them as new columns to the original data matrix, and then computing $\mathit{SSE}$ on the augmented data. Hence, using the unweighted sum approach, k-plus anticlustering leaves the original k-means criterion unchanged.

Note that a problem may arise if the weights $w_1$ and $w_2$ are ignored when optimizing the k-plus objective. Because the computation of the new attribute values $y_{ij}$ involves squaring, their range of values is expected to be very different from the raw values $x_{ij}$ (e.g., see Figure 1). This imbalance can strongly and unpredictably influence the relative weight of $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ when computing $\mathit{SSE_{kplus}}$. To ensure an equal weight of both criteria, all attributes ($x_{ij}$ and $y_{ij}$) can be standardized before computing $\mathit{SSE_{kplus}}$, e.g. via $z$-standardization.

@@

Research on multiobjective optimization indicates that solutions oftentimes fail to satisfactorily address several criteria at the same time [@ferligoj1992direct]. It is therefore unclear whether optimizing a bicriterion objective such as $\mathit{SSE_{kplus}}$ can yield a partitioning where both constituting criteria (mean and variance) are similar between groups. A motivating example illustrates that in the case of k-plus anticlustering, aiming to simultaneously optimize similarity of means and variances may be feasible. I created 14 data points following a univariate normal distribution ($M = 15$, $SD = 3$) and then generated all `r n_partitions(14, 2)` possible ways to partition the 14 data points into $K = 2$ equal-sized groups. Figure 1 shows the scatterplot of the criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ across partitions. The partition that is represented by the squared symbol has the overall best highest $\mathit{SSE}$, but is outperformed by many other partitions with regard to $\mathit{SSE_{Var}}$. Similarly, the partition that maximizes $\mathit{SSE_{Var}}$ (represented by the triangular symbol) is not well-suited to optimize $\mathit{SSE}$ at the same time. Thus, when only considering one of the objectives, the mathematically optimal partitioning fails to establish groups that are similar both with regard to their means and variances. The diamond symbol represents the partition that maximizes $\mathit{SSE_{kplus}}$, which yields `r df$kmeans[which.max(df$kplus)] / max(df$kmeans) * 100`% of the global maximum $\mathit{SSE}$, and `r df$kvar[which.max(df$kplus)] / max(df$kvar) * 100`% of the global maximum value of $\mathit{SSE_{Var}}$. Thus, using the unweighted sum approach, the optimal partition according to $\mathit{SSE_{kplus}}$ fulfills the two constituting criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ to an astonishing degree.

@@

To thoroughly evaluate the k-plus approach to anticlustering, three specific k-plus criteria were implemented in a simulation study: (i) "standard" k-plus anticlustering, minimizing differences with regard to the mean and variance; (ii) k-plus *skew/kurtosis*, minimizing differences with regard to the mean, variance, skew and kurtosis; (iii) k-plus *correlation*, minimizing differences with regard to the mean, variance and covariance structure. To implement all k-plus objectives, the unweighted sum approach was employed. Thus, to optimize a k-plus criterion, the original data set was augmented to incorporate the additional criteria and all variables---the original and the augmented variables---were $z$-standardized. Subsequently, the original k-means criterion $\mathit{SSE}$ was maximized on the basis of the augmented data set.

The three k-plus methods were compared to (i) k-means anticlustering, minimizing differences with regard to group means; (ii) diversity anticlustering, maximizing the sum of pairwise Euclidean distances within groups; (iii) a simple random allocation of elements to groups.

The simulation was implemented using the statistical programming language R [Version 4.2, @R-base]. The R package `anticlust` [Version 0.6.1, @R-anticlust; @papenberg2020] was used to optimize the anticlustering objectives. The R package `faux` [Version 1.1.0, @R-faux] was used to generate the multivariate normal data sets that were processed during the simulation. The R packages `dplyr` [Version 1.0.7, @R-dplyr], `ggplot2` [Version 3.3.5, @R-ggplot2], `papaja` [Version 0.1.1.9001, @R-papaja], `tidyr` [Version 1.1.3, @R-tidyr] and `DescTools` [Version 0.99.45, @R-desctools] were used for data analysis and result presentation. The simulation study is fully reproducible via code and data that has been made accessible via an accompanying OSF repository [@osf2023].

@@

It is maybe surprising that k-plus skew/kurtosis does not seem to have performed much better than diversity anticlustering with regard to skewness and---in particular---kurtosis, even though the method has been specifically tailored to these criteria. Figure 2 gives a more fine grained assessment of this observation: Surprisingly, for small $N$, k-plus skew/kurtosis was even worse than diversity anticlustering at minimizing $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$. However, with increasing $N$, it clearly outperformed diversity anticlustering. These results illustrate the cost of optimizing several criteria at the same time: For low $N$, k-plus skew/kurtosis was apparently preoccupied with fulfilling the objectives $\Delta \mathit{M}$ and $\Delta \mathit{SD}$. Increasing $N$ then facilitated to also address $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$.

@@

In the first example, I use norming data for the OASIS image data set that is freely available online [@kurdi2017introducing]. Kurdi et al. assembled 900 open-access color images and collected ratings on two affective dimensions: arousal and valence. @brielmann2019intense collected additional norming data by measuring how the same 900 images were rated with regard to their beauty. In my application, I used all three features; the corresponding data is available from Github.[^githubbeauty]

[^githubbeauty]: https://github.com/aenneb/OASIS-beauty

I used the function `kplus_anticlustering()` to divide the data set of 900 images into 9 groups of 100 images each. The first argument of the function call is a variable named `features`, which contains the image data as a table: 900 rows representing images and 3 columns representing the features beauty, arousal, and valence. An OSF repository accompanying this manuscript contains the full code to reproduce the example [@osf2023].

@@

kplus_anticlustering(
  features, # data table with 900 rows and 3 features
  K = 9,
  variance = TRUE,
  skew = TRUE,
  kurtosis = FALSE,
  covariances = FALSE,
  moments = NULL,
  method = "local-maximum",
  standardize = TRUE
)
```

>>>>

[^dissimilarityfromfeatures]: It is still possible to convert the attribute data into a dissimilarity matrix, e.g. by computing all pairwise Euclidean distances, and then optimize a distance-based anticlustering criterion to obtain a partitioning.

@@

Because the sample variance is basically defined as an average---and k-means anticlustering minimizes differences with regard to averages---k-means anticlustering can be employed to equalize the variance across groups. This constitutes the central insight motivating the new k-plus criterion. Thus, to obtain an adaptation of the $\mathit{SSE}$ that quantifies the degree to which the variance differs between groups, a new variable is computed for each original variable: The new variable represents the squared distance of each of the original values to the overall mean of the variable. We obtain a new data matrix $\mathbf{Y}_{N \times P} = \{y_{ij}\}_{N \times P}$ with $y_{ij} = (x_{ij} - \overline{\mathbf{x}}_j)^2$. On the basis of $\mathbf{Y}_{N \times P}$, we define the criterion $\mathit{SSE_{Var}}$, reflecting how strongly the variances of all attributes vary across the $K$ clusters:

@@

This term again has the same form as $\mathit{SSE}$ but is computed on different data. Hence, computing the k-plus criterion can be done by first computing the squared distances $y_{ij}$, appending them as new columns to the original data matrix, and then computing $\mathit{SSE}$ on the augmented data. Using the unweighted sum approach, k-plus anticlustering therefore leaves the original k-means criterion unchanged.

Note that a problem may arise if the weights $w_1$ and $w_2$ are ignored when computing $\mathit{SSE_{kplus}}$. Because the computation of the new attribute values $y_{ij}$ involves squaring, their range of values is expected to be very different from the raw values $x_{ij}$ (e.g., see Figure 1). This imbalance can strongly and unpredictably influence the relative weight of $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ when computing $\mathit{SSE_{kplus}}$. To ensure an equal weight of both criteria, all attributes ($x_{ij}$ and $y_{ij}$) can be standardized before computing $\mathit{SSE_{kplus}}$, e.g. via $z$-standardization.

@@

Research on multiobjective optimization indicates that solutions oftentimes fail to satisfactorily address several criteria at the same time [@ferligoj1992direct]. It is therefore unclear whether optimizing a bicriterion objective such as $\mathit{SSE_{kplus}}$ can yield a partitioning where both constituting criteria (mean and variance) are similar between groups. A motivating example illustrates that aiming to simultaneously optimize similarity of means and variances may be feasible. I created 14 data points following a univariate normal distribution ($M = 15$, $SD = 3$) and then generated all `r n_partitions(14, 2)` possible ways to partition the 14 data points into $K = 2$ equal-sized groups. Figure 1 shows the scatterplot of the criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ across partitions. The partition that is represented by the squared symbol has the overall best highest $\mathit{SSE}$, but is outperformed by many other partitions with regard to $\mathit{SSE_{Var}}$. Similarly, the partition that maximizes $\mathit{SSE_{Var}}$ (represented by the triangular symbol) is not well-suited to optimize $\mathit{SSE}$ at the same time. Thus, when only considering one of the objectives, the mathematically optimal partitioning fails to establish groups that are similar both with regard to their means and variances. The diamond symbol represents the partition that maximizes $\mathit{SSE_{kplus}}$, which yields `r df$kmeans[which.max(df$kplus)] / max(df$kmeans) * 100`% of the global maximum $\mathit{SSE}$, and `r df$kvar[which.max(df$kplus)] / max(df$kvar) * 100`% of the global maximum value of $\mathit{SSE_{Var}}$. Thus, using the unweighted sum approach, the optimal partition according to $\mathit{SSE_{kplus}}$ fulfills the two constituting criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ to an astonishing degree.

@@

To thoroughly evaluate the k-plus approach to anticlustering, three specific k-plus criteria were implemented in a simulation study: (i) "standard" k-plus anticlustering, minimizing differences with regard to group means and variances; (ii) k-plus *skew/kurtosis*, minimizing differences with regard to group means, variances, skewness and kurtosis; (iii) k-plus *correlation*, minimizing differences with regard to group means, variances and covariance structure. To implement all k-plus objectives, the unweighted sum approach was employed. Thus, to optimize a k-plus criterion, the original data set was augmented to incorporate the additional criteria and all variables---the original and the augmented variables---were $z$-standardized. Subsequently, the original k-means criterion $\mathit{SSE}$ was maximized on the basis of the augmented data set. The three k-plus methods were compared to (i) k-means anticlustering, minimizing differences with regard to group means, (ii) diversity anticlustering, maximizing the sum of pairwise Euclidean distances within groups, (iii) a random allocation of elements to groups.

The simulation was implemented using the statistical programming language R [Version 4.2, @R-base]. The R package `anticlust` [Version 0.6.1, @R-anticlust; @papenberg2020] was used to optimize the anticlustering objectives. The R package `faux` [Version 1.1.0, @R-faux] was used to generate the multivariate normal data sets that were processed during the simulation. The R packages `dplyr` [Version 1.0.7, @R-dplyr], `ggplot2` [Version 3.3.5, @R-ggplot2], `papaja` [Version 0.1.1.9001, @R-papaja], `tidyr` [Version 1.1.3, @R-tidyr] and `DescTools` [Version 0.99.45, @R-desctools] were used for data analysis and result presentation. The simulation study is fully reproducible via code and data that has been made accessible in an accompanying OSF repository [@osf2023].

@@

It is maybe surprising that k-plus skew/kurtosis does not seem to have performed much better than diversity anticlustering with regard to skewness and kurtosis, even though the method has been specifically tailored to these criteria. Figure 2 gives a more fine grained assessment of this observation: Surprisingly, for small $N$, k-plus skew/kurtosis was even worse than diversity anticlustering at minimizing $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$. However, with increasing $N$, it clearly outperformed diversity anticlustering. These results illustrate the cost of optimizing several criteria at the same time: For low $N$, k-plus skew/kurtosis was apparently preoccupied with fulfilling the objectives $\Delta \mathit{M}$ and $\Delta \mathit{SD}$. Increasing $N$ then facilitated to also address $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$.

@@

In the first example, I used the OASIS norming data that is freely available online [@kurdi2017introducing]. Kurdi et al. assembled 900 open-access color images and collected ratings on two affective dimensions: arousal and valence. @brielmann2019intense measured how the same 900 images were rated with regard to beauty. I used the function `kplus_anticlustering()` to divide all images into 9 groups of 100 images each. The first argument of the function call is a variable named `features`, which contains the image data as a table: 900 rows representing images and 3 columns representing the features beauty, arousal, and valence. An OSF repository accompanying this manuscript contains the full code to reproduce the example [@osf2023].

@@

kplus_anticlustering(
  features,
  K = 9,
  variance = TRUE,
  skew = TRUE,
  kurtosis = FALSE,
  covariances = FALSE,
  moments = NULL,
  method = "local-maximum",
  standardize = TRUE
)
```
